---
title: "2023-10-16-NCM"
author: "kim soyeon"
date: "2023-10-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(phyloseq)
library(devtools)
library(tibble)
library(ape)
library(vegan)
library(FSA)
library(eulerr)
library(grid)e
library(gridExtra)

ps <- readRDS("./ps.rds")

ps.g <- subset_samples(ps, body.site %in% c("gut"))
ps.t <- subset_samples(ps, body.site %in% c("tongue"))


```


```{r}

# spp = t(otu_table(ps.gt)); pool=NULL; stats=TRUE; taxon= tax_table(ps.gt)

```



```{r}

#spp: A community table for communities of interest with local communities/samples as rows and taxa as columns. All samples must be rarefied to the same depth.
#pool: A community table for defining source community (optional; Default=NULL).
#taxon: A table listing the taxonomic calls for each otu, with OTU ids as row names and taxonomic classifications as columns.
#If stats=TRUE the function will return fitting statistics.
#If stats=FALSE the function will return a table of observed and predicted values for each otu.

```

### Import data

```{r, message=FALSE, warning=FALSE}
phyloseq = ps.g
meta <- sample_data(phyloseq)
meta$X.Sample <- rownames(meta)
sample_data(phyloseq) <- meta

## Check how many reads you have in each of the samples. This will tell you if you need to re-do anything
# Get read counts and make a new dataframe with this data
read_count = data.frame("count" = colSums(otu_table(phyloseq))) %>%
  rownames_to_column(var="X.Sample") %>%
  inner_join(data.frame(sample_data(phyloseq)), by="X.Sample") %>%
  arrange(-count) %>%
  mutate(X.Sample=factor(X.Sample, levels=X.Sample))

# Now plot read count for each sample. The horizontal line represents a 2000 read threshold
ggplot(data=read_count, aes(x=X.Sample, y=log10(count), fill=body.site)) +
  geom_bar(stat="identity") +
  labs(x="Sample", y="Log10(Read count)") +
  geom_hline(yintercept=log10(10000)) +
  geom_hline(yintercept=log10(2000)) +
  theme(text = element_text(size=16),
        axis.text.x = element_blank())


```



relative abundance rather than readcounts.

```{r, message=FALSE, warning=FALSE}
# Rarefy to an even depth
set.seed(42)  # setting seed for reproducibility
phyloseq.rare = rarefy_even_depth(phyloseq)

# Normalize read counts (this gives relative abundance)
phyloseq.norm = transform_sample_counts(phyloseq.rare, function(x) x/sum(x))

```

## Testing neutral model fit

Now lets fit the sloan neutral model to our data. I will do this for the entire dataset together (irregardless of land use). This code was adapted from Adam Burns et al. 2016.

Burns, A., Stephens, W., Stagaman, K. et al. Contribution of neutral processes to the assembly of gut microbial communities in the zebrafish over host development. ISME J 10, 655–664 (2016) doi:10.1038/ismej.2015.142

```{r, message=FALSE, warning=FALSE}

# Extract the OTU table from pthe phyloseq object
OTU.table = t(otu_table(phyloseq.rare))

# Calculate the number of individuals in the meta community (Average read depth)
N <- mean(apply(OTU.table, 1, sum))

# Calculate the average relative abundance of each taxa across communities
p.m <- apply(OTU.table, 2, mean)
p.m <- p.m[p.m != 0]
p <- p.m/N
p.df = data.frame(p) %>%
  rownames_to_column(var="OTU")

# Calculate the occurrence frequency of each taxa
OTU.table.bi <- 1*(OTU.table>0)
freq.table <- apply(OTU.table.bi, 2, mean)
freq.table <- freq.table[freq.table != 0]
freq.df = data.frame(OTU=names(freq.table), freq=freq.table)

#Combine
C <- inner_join(p.df,freq.df, by="OTU") %>%
  arrange(p)
# Remove rows with any zero (absent in either source pool or local communities). You already did this, but just to make sure we will do it again.
C.no0 <- C %>%
  filter(freq != 0, p != 0)

#Calculate the limit of detection
d <- 1/N

##Fit model parameter m (or Nm) using Non-linear least squares (NLS)
p.list <- C.no0$p
freq.list <- C.no0$freq
m.fit <- nlsLM(freq.list ~ pbeta(d, N*m*p.list, N*m*(1-p.list), lower.tail=FALSE), start=list(m=0.1))
m.ci <- confint(m.fit, 'm', level=0.95)
m.sum <- summary(m.fit)
m.coef = coef(m.fit)

freq.pred <- pbeta(d, N*coef(m.fit)*p.list, N*coef(m.fit)*(1-p.list), lower.tail=FALSE)
Rsqr <- 1 - (sum((freq.list - freq.pred)^2))/(sum((freq.list - mean(freq.list))^2))

# Get table of model fit stats
fitstats <- data.frame(m=m.coef, m.low.ci=m.ci[1], m.up.ci=m.ci[2], 
                       Rsqr=Rsqr, p.value=m.sum$parameters[4], N=N, 
                       Samples=nrow(OTU.table), Richness=length(p.list), 
                       Detect=d)

# Get confidence interval for predictions
freq.pred.ci <- binconf(freq.pred*nrow(OTU.table), nrow(OTU.table), alpha=0.05, method="wilson", return.df=TRUE)

# Get table of predictions
pred.df <- data.frame(metacomm_RA=p.list, frequency=freq.pred, 
                      frequency_lowerCI=freq.pred.ci[,2], 
                      frequency_upperCI=freq.pred.ci[,3]) %>%
  unique()

# Get table of observed occupancy and abundance
obs.df = C.no0 %>%
  rename(metacomm_RA = p, frequency=freq)

obs.df <-  obs.df %>%
  left_join(pred.df, by = "metacomm_RA") %>%
  mutate(New_Column = case_when(
    frequency.x >= frequency_lowerCI & frequency.x <= frequency_upperCI ~ "fit",
    frequency.x > frequency_upperCI ~ "upper",
    frequency.x < frequency_lowerCI ~ "lower",
    TRUE ~ "NA"
  )) %>%
  select(-frequency.y, -frequency_lowerCI, -frequency_upperCI)

colnames(obs.df)[3] <- "frequency"



```

Plot the model and observed occupancy and metacommunity abundance values. 

```{r}  

fulldata.model.plot = 
  ggplot(data=obs.df) +
    geom_point(data=obs.df, aes(x=log10(metacomm_RA), y=frequency, color = New_Column), 
               alpha=.2, size=2) +
    geom_line(data=pred.df, aes(x=log10(metacomm_RA), y=frequency), color="black") + 
    geom_line(data=pred.df, aes(x=log10(metacomm_RA), y=frequency_lowerCI), linetype=2, color="black") + 
    geom_line(data=pred.df, aes(x=log10(metacomm_RA), y=frequency_upperCI), linetype=2, color="black") + 
    geom_text(data=fitstats, aes(label = paste("R^2 == ", round(Rsqr, 3))), 
              x=-3.8, y=0.95, size=4, parse=TRUE) +
    geom_text(data=fitstats, aes(label = paste("italic(m) ==", round(m, 3))), 
              x=-3.8, y=0.85, size=4, parse=TRUE) + 
    labs(x="Log10 abundance in\nmetacommunity", y="Frequency detected") +
    theme_bw() +
    theme(axis.line = element_line(color="black"),
          legend.position = "none",
          axis.title = element_text(size=14),
          axis.text = element_text(size=12))

fulldata.model.plot

ggsave("region_neutral_model.png", plot=fulldata.model.plot, device="png",
      path="./",
      width=10, height=10, units="cm")
```
	
	
	
```{r}


Check_read <- function(phyloseq){
  meta <- sample_data(phyloseq)
  meta$SampleID <- rownames(meta)
  sample_data(phyloseq) <- meta
  
  ## Check how many reads you have in each of the samples. This will tell you if you need to re-do anything
  # Get read counts and make a new dataframe with this data
  read_count = data.frame("count" = colSums(otu_table(phyloseq))) %>%
    rownames_to_column(var="SampleID") %>%
    inner_join(data.frame(sample_data(phyloseq)), by="SampleID") %>%
    arrange(-count) %>%
    mutate(SampleID=factor(SampleID, levels=SampleID))
  
  print(read_count)
  # Now plot read count for each sample. The horizontal line represents a 2000 read threshold
  ggplot(data=read_count, aes(x=SampleID, y=log10(count), fill=body.site)) +
    geom_bar(stat="identity") +
    labs(x="Sample", y="Log10(Read count)") +
    geom_hline(yintercept=log10(10000)) +
    geom_hline(yintercept=log10(2000)) +
    theme(text = element_text(size=16),
          axis.text.x = element_blank())
  

  
}
  
NCM_fit <- function(phyloseq, rarefy = "min"){
  
  if (rarefy == "min") {
    # Rarefy to an even depth
    set.seed(42)  # setting seed for reproducibility
    phyloseq.rare = rarefy_even_depth(phyloseq)
    } else {
      phyloseq.rare = phyloseq
      }
  # Normalize read counts (this gives relative abundance)
  phyloseq.norm = transform_sample_counts(phyloseq.rare, function(x) x/sum(x))
  
  
  # Extract the OTU table from pthe phyloseq object
  OTU.table = t(otu_table(phyloseq.rare))
  
  # Calculate the number of individuals in the meta community (Average read depth)
  N <- mean(apply(OTU.table, 1, sum))
  
  # Calculate the average relative abundance of each taxa across communities
  p.m <- apply(OTU.table, 2, mean)
  p.m <- p.m[p.m != 0]
  p <- p.m/N
  p.df = data.frame(p) %>%
    rownames_to_column(var="OTU")
  
  # Calculate the occurrence frequency of each taxa
  OTU.table.bi <- 1*(OTU.table>0)
  freq.table <- apply(OTU.table.bi, 2, mean)
  freq.table <- freq.table[freq.table != 0]
  freq.df = data.frame(OTU=names(freq.table), freq=freq.table)
  
  #Combine
  C <- inner_join(p.df,freq.df, by="OTU") %>%
    arrange(p)
  # Remove rows with any zero (absent in either source pool or local communities). You already did this, but just to make sure we will do it again.
  C.no0 <- C %>%
    filter(freq != 0, p != 0)
  
  #Calculate the limit of detection
  d <- 1/N
  
  ##Fit model parameter m (or Nm) using Non-linear least squares (NLS)
  p.list <- C.no0$p
  freq.list <- C.no0$freq
  m.fit <- nlsLM(freq.list ~ pbeta(d, N*m*p.list, N*m*(1-p.list), lower.tail=FALSE), start=list(m=0.1))
  m.ci <- confint(m.fit, 'm', level=0.95)
  m.sum <- summary(m.fit)
  m.coef = coef(m.fit)
  
  freq.pred <- pbeta(d, N*coef(m.fit)*p.list, N*coef(m.fit)*(1-p.list), lower.tail=FALSE)
  Rsqr <- 1 - (sum((freq.list - freq.pred)^2))/(sum((freq.list - mean(freq.list))^2))
  
  # Get table of model fit stats
  fitstats <- data.frame(m=m.coef, m.low.ci=m.ci[1], m.up.ci=m.ci[2], 
                         Rsqr=Rsqr, p.value=m.sum$parameters[4], N=N, 
                         Samples=nrow(OTU.table), Richness=length(p.list), 
                         Detect=d)
  
  # Get confidence interval for predictions
  freq.pred.ci <- binconf(freq.pred*nrow(OTU.table), nrow(OTU.table), alpha=0.05, method="wilson", return.df=TRUE)
  
  # Get table of predictions
  pred.df <- data.frame(metacomm_RA=p.list, frequency=freq.pred, 
                        frequency_lowerCI=freq.pred.ci[,2], 
                        frequency_upperCI=freq.pred.ci[,3]) %>%
    unique()
  
  # Get table of observed occupancy and abundance
  obs.df = C.no0 %>%
    rename(metacomm_RA = p, frequency=freq)
  
  obs.df <-  obs.df %>%
    left_join(pred.df, by = "metacomm_RA") %>%
    mutate(New_Column = case_when(
      frequency.x >= frequency_lowerCI & frequency.x <= frequency_upperCI ~ "fit",
      frequency.x > frequency_upperCI ~ "upper",
      frequency.x < frequency_lowerCI ~ "lower",
      TRUE ~ "NA"
    )) %>%
    select(-frequency.y, -frequency_lowerCI, -frequency_upperCI)
  
  colnames(obs.df)[3] <- "frequency"
  
  
  
  fulldata.model.plot = 
    ggplot(data=obs.df) +
      geom_point(data=obs.df, aes(x=log10(metacomm_RA), y=frequency, color = New_Column), size=2, alpha = 0.8) +
      geom_line(data=pred.df, aes(x=log10(metacomm_RA), y=frequency), color="black") + 
      geom_line(data=pred.df, aes(x=log10(metacomm_RA), y=frequency_lowerCI), linetype=2, color="black") + 
      geom_line(data=pred.df, aes(x=log10(metacomm_RA), y=frequency_upperCI), linetype=2, color="black") + 
      geom_text(data=fitstats, aes(label = paste("R^2 == ", round(Rsqr, 3))), 
                x=-3.8, y=0.95, size=4, parse=TRUE) +
      geom_text(data=fitstats, aes(label = paste("italic(m) ==", round(m, 3))), 
                x=-3.8, y=0.85, size=4, parse=TRUE) + 
      labs(x="Log10 abundance in\nmetacommunity", y="Frequency detected") +
      theme_bw() +
      theme(axis.line = element_line(color="black"),
            legend.position = "none",
            axis.title = element_text(size=14),
            axis.text = element_text(size=12))
  
  return(fulldata.model.plot)

}



```

```{r}
Check_read(ps.g)
p.gut <- NCM_fit(ps.g)

Check_read(ps.t)
p.tng <- NCM_fit(ps.t)

```
